Bird beak's elaboration and exploration

### 2021-09-14 Analysis/03-elaboration_exploration_analyses.Rmd
Update 03-elaboration_exploration_analyses.Rmd

All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
Although `beer` is intentionally designed to be portable and shareable, we advise workers to use it with moderation and tailor its consumption to their specific research needs.
Some analyses in `beer` do take some time to run so we advise users not to operate heavy machinery while using it.
Don't use it while driving.
The following code snippets are used to illustrate the implementation of the functions.
See <!-- §§§run_replicates_script --> for running the actual analyses.

@@

```{r, eval = TRUE, echo = FALSE}
library(beer)
## Loading the MCMCglmm
load("../Data/processed/combined_results.rda")
## Loading the tree and data
data(demo_data)
data(demo_tree)
## Setting the colour vector
colour_vector <- c("orange", "blue", "darkgreen", "grey")
```

@@

To analyse the exploration and elaboration aspects of both clades and tips, we first selected 1000 random covariance matrices from the combined MCMCglmm posterior variance-covariance matrices.

```{r, eval = TRUE}
covar_matrices <- MCMCglmm.dispRity(
                  data          = demo_data,
                  posteriors    = combined_results,
                  group         = MCMCglmm.levels(combined_results)[1:4],
                  rename.groups = c("gulls", "plovers", "sandpipers", "phylogeny"))
```

```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5}
## Plotting just the space
sauron.plot(covar_matrices,
            col = colour_vector,
            main = "demo space",
            legend = TRUE)
```

We use these covariance matrices as the basis of the phylogenetic main axes of variation.
These axes correspond to the main axes of variation in the $N_DIMENSIONS$ trait space at different phylogenetic levels:

@@

```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5}
get.one.axis <- function(data, axis = 1, level = 0.95, dimensions) {

@@

Figure: the main axis of variation (in orange) of the 95% CI ellipse of the variance-covariance matrix (in blue) of 20 elements in 2 dimensions (in black).

In practice, the coordinates of these axes end points were calculated directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigenvalues and the variance-covariance matrix's eigenvectors.

## Major axes calculation
<!-- % NC: Mathematically seems sound, but I'm no expert. But very hard to parse for anyone who hasn't been looking at this in detail. Think fo the poor palaeontologists!
 -->
For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and a radius composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:

@@

(in other words $O_{n}$ is the matrix representing each edge point of the unit hypersphere of $n$ dimensions and of radius $1$);

@@

We can get $M_{n}$ matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scale 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:

@@

```{r, fig.height = 6, fig.width = 6}
sauron.plot(covar_matrices,
            col = colour_vector,
            main = "Average ellipses and\nmajor axes",
            major.axes = mean,
            ellipses = mean,
            legend = TRUE,
            cex = 0.5)
```

```{r, fig.height = 12, fig.width = 6}
par(mfrow = c(2,1))
## Plotting 100 axes
sauron.plot(covar_matrices,
            n = 100,
            col = colour_vector,
            main = "100 random major axes",
            major.axes = TRUE,
            legend = TRUE,
            points = FALSE,
            xlim = c(-2, 2),
            ylim = c(-2, 2))
## Plotting 100 ellipses
sauron.plot(covar_matrices,
            n = 100,
            col = colour_vector,
            main = "100 random ellipses",
            ellipses = TRUE,
            legend = TRUE,
            points = FALSE,
            xlim = c(-2, 2),
            ylim = c(-2, 2))
```

#### Measuring projection and rejection

@@

#### Generalisation of projection onto any vector in a set space

@@

Following the ways to: <!-- %NC: Not sure what you mean here -->
 1. estimate the variance-covariance matrices of the phylogenetic effects in trait space (02-MCMCglmm_mini_chains.Rmd);
 2. extracting the major axes of variation of these VCV matrices for the different levels in the phylogeny;
 3. and calculating the elaboration and exploration scores for elements in the trait space based on these major axes.

We can measure these scores for each clade and each species separately.

To do so, for a randomly selected posterior variance-covariance matrix we defined $\vec{a}$ as the major axes for the clade of interest and projected it onto $\vec{B}$ as the major overall phylogenetic effect (see above).
For each pair of $\vec{a}$ and $\vec{B}$, we translated the origin of $\vec{a}$ onto the origin of $\vec{B}$ and measured the projection (elaboration), rejection (exploration) and angle (degree of exploration).

```{r, fig.height = 6, fig.width = 18}
## Projecting each group's major axes on the phylogenetic major axes
group_elaboration <- dispRity.covar(covar_matrices,
               metric = projections.covar,
               between.groups = list(c(1,4), c(2,4), c(3,4)),
               measure = "position")
group_exploration <- dispRity.covar(covar_matrices,
               metric = projections.covar,
               between.groups = list(c(1,4), c(2,4), c(3,4)),
               measure = "distance")
group_angle <- dispRity.covar(covar_matrices,
               metric = projections.covar,
               between.groups = list(c(1,4), c(2,4), c(3,4)),
               measure = "degree")
par(mfrow = c(1,3))
plot(group_elaboration, ylab = "elaboration", col = colour_vector[-4])
plot(group_exploration, ylab = "exploration", col = colour_vector[-4])
plot(group_angle, ylab = "angle", col = colour_vector[-4])
```

The results are reported as boxplots for each group indicating which group explores/elaborates more (and to which degree - angle) relatively to each other.

Similarly to the analyses described above, we ran the analyses for each species but here we directly compared each species position in the space to the same vector $\vec{B}$ described above and reported the results for each groups.

```{r, eval = FALSE}
## Get the mean axes for each level
major_axes <- axis.covar(covar_matrices, sample = mean)

## Projecting each group's major axes on the phylogenetic major axes
phy_group_elaboration <- dispRity(covar_matrices,
               dimensions = covar_matrices$call$dimensions,
               metric  = dispRity::projections,
               measure = "position",
               point1  = major_axes$phylogeny[[1]][1,],
               point2  = major_axes$phylogeny[[1]][2,])
phy_group_exploration <- dispRity(covar_matrices,
               dimensions = covar_matrices$call$dimensions,
               metric  = dispRity::projections,
               measure = "distance",
               point1  = major_axes$phylogeny[[1]][1,],
               point2  = major_axes$phylogeny[[1]][2,])
phy_group_angle <- dispRity(covar_matrices,
               dimensions = covar_matrices$call$dimensions,
               metric  = dispRity::projections,
               measure = "degree",
               point1  = major_axes$phylogeny[[1]][1,],
               point2  = major_axes$phylogeny[[1]][2,])

## TODO: add option in dispRity plot to only select some groups
par(mfrow = c(1,3))
plot(phy_group_elaboration, col = colour_vector)
plot(phy_group_exploration, ylab = "exploration", col = colour_vector)
plot(phy_group_angle, ylab = "angle", col = colour_vector)

## Global analyses
tip_results_global <- analyses.tip(data = demo_data[, c(1:4)],
                                   axes = major_axes$animal)
```

And here are the values plotted by group:

```{r, eval = FALSE}
## Plot results
plot.analyses.tip(tip_results_global,
                  col = colour_vector,
                  group = split(rownames(demo_data), f = demo_data$clade),
                  main = "Tips vs. whole phylogeny")
```

Note that here only the tip positions are projected on the global phylogenetic axes.

Finally, we also looked at the exploration elaboration profiles within each clade, by replacing the projection of the elements onto $\vec{B}$ for each clade to the projection onto their respective clade major axes $\vec{a}$.

```{r, eval = FALSE}
## Results within groups
tip_results_within <- analyses.tip(
                    data = demo_data[, c(1:4)],
                    axes = major_axes[-length(major_axes)],
                    group = split(rownames(demo_data), f = demo_data$clade))
```

```{r, eval = FALSE}
## Plot results
plot.analyses.tip(tip_results_within,
                  col = colour_vector,
                  group = split(rownames(demo_data), f = demo_data$clade),
                  main = "Tips vs. clade")
```

But this is probably best to look at in a species by species fashion:

TODO

>>>>

<!-- TG: The analyses will be eventually run in dispRity so we will miss out the following notices: 
Some analyses in `beer` do take some time to run so we advise users not to operate heavy machinery while using it.
Don't use it while driving. -->
The following code snippets are used to illustrate the implementation of the functions.

@@

```{r, eval = TRUE, echo = FALSE, print = FALSE}
library(beer)
## Loading the MCMCglmm
load("../Data/processed/combined_results.rda")
## Loading the tree and data
data(demo_data)
data(demo_tree)
## Setting the colour vector
colour_vector <- c("orange", "blue", "darkgreen", "grey")
```

@@

To analyse the exploration and elaboration aspects of both clades and tips, we first selected 1000 random covariance matrices from the combined MCMCglmm posterior variance-covariance matrices represented in this space.

```{r, eval = TRUE, echo = FALSE}
covar_matrices <- MCMCglmm.dispRity(
                  data          = demo_data,
                  posteriors    = combined_results,
                  group         = MCMCglmm.levels(combined_results)[1:4],
                  rename.groups = c("gulls", "plovers", "sandpipers", "phylogeny"))
```

```{r, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5, echo = FALSE}
## Plotting just the space
sauron.plot(covar_matrices,
            col = colour_vector,
            main = "demo space",
            legend = TRUE)
```

From this space and the mini chains run before, we can extract the posterior variance covariance matrices to calculate the major axes of variation in the `r dim(length(covar_matrices$call$dimensions))` dimensions trait space at different levels:

@@

```{r, echp = FALSE, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5, echo = FALSE}
get.one.axis <- function(data, axis = 1, level = 0.95, dimensions) {

@@

Figure: the main axis of variation (in orange) of the 95% CI ellipse of the variance-covariance matrix (in blue) of 20 elements in 2 dimensions (in black).
Note that when estimated using a `mcmcmcglmmm` approach there are as many 95% CI ellipses and corresponding main axes as they are posterior samples.

In practice, we calculated the coordinates of these axes end points  directly from the cross product between the 95% unit hypersphere scaled by the variance-covariance matrix's squared eigenvalues and the variance-covariance matrix's eigenvectors (see details 1).

```{r, echo = FALSE, fig.height = 6, fig.width = 6}
sauron.plot(covar_matrices,
            col = colour_vector,
            main = "Average ellipses and\nmajor axes",
            major.axes = mean,
            ellipses = mean,
            legend = TRUE,
            cex = 0.5)
```

```{r, echo = FALSE, fig.height = 12, fig.width = 6}
par(mfrow = c(2,1))
## Plotting 100 axes
sauron.plot(covar_matrices,
            n = 100,
            col = colour_vector,
            main = "100 random major axes",
            major.axes = TRUE,
            legend = TRUE,
            points = FALSE,
            xlim = c(-2, 2),
            ylim = c(-2, 2))
## Plotting 100 ellipses
sauron.plot(covar_matrices,
            n = 100,
            col = colour_vector,
            main = "100 random ellipses",
            ellipses = TRUE,
            legend = TRUE,
            points = FALSE,
            xlim = c(-2, 2),
            ylim = c(-2, 2))
```

We measured the exploration and elaboration using two main approaches:
 * By clades (i.e exploration/elaboration **between** clades): where we calculated the projections of the major axes of each clade onto the global phylogenetic major axes.
 * By species (i.e exploration/elaboration **within** clades): where we calculated the projections of each species onto 1) the global phylogenetic major axes and 2) the phylogenetic major axes of their respective clade.

For each posterior of the `mcmcmcglmmm` model, we projected each group's major axes onto the global phylogenetic major axis of the specific posterior sample.
In practice, for the projection to work, we translated and centred each group's major axes onto the origin of the global phylogenetic major axis (so that both origins have the coordinates 0) and then measure the projection and rejection as specified in Details 2.

```{r, fig.height = 6, fig.width = 18, echo = FALSE}
## Run the analyses between groups
results_between <- covar.projections.wrapper(covar_matrices,
                                             type = "between",
                                             base = "phylogeny",
                                             dispRity.out = TRUE)
## Plot the results
wrap.plot <- function(results, col, main) {
    ## How many sub plots?
    n <- length(results)
    ## Which will be the centre plot?
    centre <- ceiling(n/2)
    ## Plot everything
    par(mfrow = c(1, n))
    for(i in 1:n) {
        plot(results[[i]], col = col, main = ifelse(i == centre, main, ""))
    }
}
wrap.plot(results_between,
          col = colour_vector[-4],
          main = "Groups difference")
```
Figure: these results show the projections ("position"), rejection ("distance") and angles ("degree") of the posteriors of each group's major axes onto the corresponding global phylogenetic major axes. For each panel, the highest the score the more the clade is elaborating, exploring, exploring perpendicular to the main axes for the position, distance and degree respectively.

Similarly to the analyses described above, we ran the analyses for each species.
Here, instead of projecting the major axes of the groups onto the global phylogenetic major axis we projected each species onto the phylogenetic major axis (for each posterior distribution).

```{r, fig.height = 6, fig.width = 18, echo = FALSE, eval = TRUE}
## Run the analyses of groups on the phylogeny
results_phylo <- covar.projections.wrapper(covar_matrices,
                                            type = "within",
                                            base = "phylogeny",
                                            dispRity.out = TRUE)
## Plot the results
wrap.plot(results_phylo,
          col = colour_vector[-4],
          main = "Projection on the\nglobal phylogeny axes")
```
Figure: the projection of each species per group onto the posterior distribution of global major phylogenetic axis.

Finally, we ran the same analyses but instead of the projection of the species onto the global phylogenetic major axes, we projected them onto their respective group's phylogenetic major axes.

```{r, fig.height = 6, fig.width = 18, echo = FALSE, eval = TRUE}
## Run the analyses groups on themselves
results_within <- covar.projections.wrapper(covar_matrices,
                                            type = "within",
                                            dispRity.out = TRUE)
## Plot the results
wrap.plot(results_within,
          col = colour_vector,
          main = "Projection on the\ngroup's phylogeny axes")
```
Figure: the projection of each species per group onto the posterior distribution of each clade's major phylogenetic axis (the grey distributions corresponds to the projections of all the species, irregardless of their group, onto the global phylogenetic major axes).

>TG: TODO: do this plot on a phylogeny

# Supplementary details

## Details 1: Major axes calculation

For $O_{n}$, the unit hypersphere matrix of _n_ dimensions and a radius composed of the two identity matrices $I_{n}$ and $-I_{n}$ so that:

@@

In other words $O_{n}$ is the matrix representing each edge point of the unit hypersphere of $n$ dimensions and of radius $1$;

@@

We can get $M_{n}$ matrix containing all the edge coordinates of the 0.95 CI hypersphere from $VCV_{n}$ using the transposition of the cross product between the eigenvectors **v** and the product of the scaled 0.95 CI unit sphere $O'_{n}$ and the eigenvalues $\lambda$:

@@

## Details 2: Measuring projection and rejection

@@

### Generalisation of projection onto any vector in a set space

@@
